---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# qrme

`qrme` is the companion R package for

> Callaway, B., Li, T., Murtazashvili, I., and Tsyawo, E. S. (2024).
> *Distributional Effects with Two-Sided Measurement Error: An Application
> to Intergenerational Income Mobility.*
> [arXiv:2107.09235](https://arxiv.org/abs/2107.09235)

It provides two main estimators:

- **`qrme()`** — Quantile regression with measurement error in the outcome
  variable. This is an EM-based implementation of the estimator developed by
  [Hausman, Liu, Luo, and Palmer (2021, *Econometrica*)](https://doi.org/10.3982/ECTA14563),
  extended to support finite Gaussian mixtures for the measurement error
  distribution. The pseudo-EM algorithm alternates between a Metropolis–Hastings
  MCMC step (E-step) to integrate out the latent outcome and a standard quantile
  regression M-step, iterating until convergence. `qrme()` is a useful
  standalone tool for any setting where the dependent variable is measured with
  classical error.

- **`tsme()`** — Two-sided measurement error: joint quantile regression when
  *both* the outcome and a continuous regressor are measured with error.
  Combines two `qrme()` fits via a copula to recover the joint distribution
  and a range of distributional parameters such as quantile-on-quantile
  transition matrices and poverty measures.

## Installation

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("bcallaway11/qrme")
```

## Usage

### Quantile regression with measurement error in the outcome (`qrme`)

```{r, eval = FALSE}
library(qrme)

# Simulate data: Y* = 1 + X + u, Y = Y* + e (measurement error)
set.seed(42)
n  <- 500
X  <- runif(n)
Ys <- 1 + X + rnorm(n)         # latent outcome
Y  <- Ys + rnorm(n, sd = 0.5)  # observed outcome with ME
dd <- data.frame(Y = Y, X = X)

fit <- qrme(
  formula = Y ~ X,
  data    = dd,
  tau     = c(0.25, 0.5, 0.75)
)

# Coefficient matrix: one row per quantile
fit$bet
```

Key arguments:

| Argument | Default | Description |
|---|---|---|
| `tau` | `0.5` | Quantile(s) to estimate |
| `nmix` | `1` | Number of Gaussian mixture components for ME distribution |
| `conv_crit` | `"params"` | Convergence criterion: `"params"` or `"loglik"` |
| `tol` | `1e-2` | Convergence tolerance |
| `maxit` | `100` | Maximum EM iterations |
| `mcmc_draws` | `200` | MCMC draws per EM E-step |
| `mcmc_burnin` | `100` | MCMC burn-in draws |
| `se` | `FALSE` | Compute bootstrap standard errors |
| `verbose` | `FALSE` | Print progress (numeric levels 0–3) |

### Two-sided measurement error (`tsme`)

```{r, eval = FALSE}
res <- tsme(
  data     = dd,
  Yformla  = Y ~ X,
  Tformla  = T ~ X,
  tau      = c(0.25, 0.5, 0.75),
  tvals    = mean(dd$T),
  povline  = quantile(dd$Y, 0.2)
)

# Transition matrix (distributional mobility summary)
res$meTmat
```

`tsme()` fits a `qrme()` model for each variable, then links them with a
copula to estimate joint distributions and mobility parameters such as
quantile-on-quantile transition matrices and poverty measures.

## References

- Callaway, B., Li, T., Murtazashvili, I., and Tsyawo, E. S. (2024).
  Distributional Effects with Two-Sided Measurement Error: An Application to
  Intergenerational Income Mobility. arXiv:2107.09235.

- Hausman, J., Liu, H., Luo, Y. and Palmer, C. (2021). Errors in the
  dependent variable of quantile regression models. *Econometrica*, 89(2),
  pp. 849–873.
