---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# qrme

`qrme` is the companion R package for

> Callaway, B., Li, T., Murtazashvili, I., and Tsyawo, E. S. (2026).
> *Distributional Effects with Two-Sided Measurement Error: An Application
> to Intergenerational Income Mobility.*
> [arXiv:2107.09235](https://arxiv.org/abs/2107.09235)

It provides two main estimators:

- **`qrme()`** — Quantile regression with measurement error in the outcome
  variable. This is an EM-based implementation of the estimator developed by
  [Hausman, Liu, Luo, and Palmer (2021, *Econometrica*)](https://doi.org/10.3982/ECTA14563),
  extended to support finite Gaussian mixtures for the measurement error
  distribution. The pseudo-EM algorithm alternates between a Metropolis–Hastings
  MCMC step (E-step) to integrate out the latent outcome and a standard quantile
  regression M-step, iterating until convergence. `qrme()` is a useful
  standalone tool for any setting where the dependent variable is measured with
  classical error.

- **`tsme()`** — Two-sided measurement error: joint quantile regression when
  *both* the outcome and a continuous regressor are measured with error.
  Combines two `qrme()` fits via a copula to recover the joint distribution
  and a range of distributional parameters such as quantile-on-quantile
  transition matrices and poverty measures.

## Installation

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("bcallaway11/qrme")
```

## Usage

### Quantile regression with measurement error in the outcome (`qrme`)

```{r, eval = FALSE}
library(qrme)

# Simulate data: Y* = 1 + X + u, Y = Y* + e (measurement error)
set.seed(42)
n  <- 500
X  <- runif(n)
Ys <- 1 + X + rnorm(n)         # latent outcome
Y  <- Ys + rnorm(n, sd = 0.5)  # observed outcome with ME
dd <- data.frame(Y = Y, X = X)

fit <- qrme(
  formula = Y ~ X,
  data    = dd,
  tau     = c(0.25, 0.5, 0.75)
)

# Coefficient matrix: one row per quantile
fit$bet
```

Key arguments:

| Argument | Default | Description |
|---|---|---|
| `tau` | `0.5` | Quantile(s) to estimate |
| `n_mix` | `1` | Number of Gaussian mixture components for ME distribution |
| `conv_criterion` | `"params"` | Convergence criterion: `"params"` or `"loglik"` |
| `tol` | `1e-2` | Convergence tolerance |
| `max_em_iters` | `100` | Maximum EM iterations |
| `mcmc_draws` | `200` | MCMC draws per EM E-step |
| `mcmc_burn_in` | `100` | MCMC burn-in draws |
| `se` | `FALSE` | Compute bootstrap standard errors |
| `verbose` | `FALSE` | Print progress (numeric levels 0–3) |

### Two-sided measurement error (`tsme`)

```{r, eval = FALSE}
res <- tsme(
  data     = dd,
  y_formula  = Y ~ X,
  t_formula  = T ~ X,
  tau      = c(0.25, 0.5, 0.75),
  t_values    = mean(dd$T),
  pov_line  = quantile(dd$Y, 0.2)
)

# Transition matrix (distributional mobility summary)
res$me_tmat
```

`tsme()` fits a `qrme()` model for each variable, then links them with a
copula to estimate joint distributions and mobility parameters such as
quantile-on-quantile transition matrices and poverty measures.

## Tuning parameters

The EM algorithm involves several parameters with different roles. The table
below groups them by the nature of their tradeoff.

| Parameter | Default | Role |
|---|---|---|
| `tol` | `1e-2` | Pure speed–accuracy: loosen to converge faster, tighten for more precise estimates |
| `max_em_iters` | `100` | Hard iteration cap; does not affect per-iteration cost |
| `conv_patience` | `1` | Consecutive iterations below `tol` required; raise to 2 if using `conv_criterion = "loglik"` |
| `mcmc_draws` | `200` | **Fundamental tradeoff**: more draws = better E-step but slower per iteration |
| `mcmc_burn_in` | `100` | Burn-in fraction of `mcmc_draws`; 30–50% is typical |
| `loglik_draws` | `100` | MC draws for log-likelihood convergence check; ignored when `conv_criterion = "params"` |
| `proposal_sd` | adaptive | MH step size; auto-updated each iteration to track the current ME scale |
| `n_mix` | `1` | Number of mixture components — a model choice, not a tuning parameter |
| `start_beta` | naive QR | Defaults to QR ignoring ME, which is consistent under symmetric error |
| `start_mu` | data-informed | Evenly spaced over ±0.25 sd(Y) for n_mix > 1; 0 for n_mix = 1 |
| `start_sigma` | data-informed | Backed out so the mixture SD equals 0.5 sd(Y) |
| `start_pi` | uniform | Equal weights across components |

**`mcmc_draws` and `mcmc_burn_in`** are the most consequential parameters.
Too few effective draws (= `mcmc_draws − mcmc_burn_in`) per observation produces
a noisy E-step, which can cause the outer EM to wander or fail to converge —
this is a correctness problem, not just a speed problem. The defaults of 200/100
work well in most settings; reduce only for exploratory runs.

**`proposal_sd`** controls the Metropolis–Hastings step size and determines the
effective sample size of the MCMC chain. By default it is initialised from the
starting ME parameters and updated each EM iteration to track the current ME
distribution scale. A healthy acceptance rate is roughly 20–70%; the acceptance
rate is printed at `verbose = 2` and a warning is issued if the final-iteration
rate falls outside 10–90%. If the warning fires, pass `proposal_sd` explicitly
to override the adaptive default.

**`n_mix`** is selected by model fit rather than tuned. Use `logLik()`, `AIC()`,
and `BIC()` on fitted objects to compare across values; `n_mix = 1` is the right
default for classical additive Gaussian measurement error.

## Learn more

- **Vignettes** — [Introduction to `qrme()`](https://bcallaway11.github.io/qrme/articles/qrme-intro.html) |
  [NLSY97 application with `tsme()`](https://bcallaway11.github.io/qrme/articles/tsme-application.html)
- **Package website** — <https://bcallaway11.github.io/qrme>

## References

- Callaway, B., Li, T., Murtazashvili, I., and Tsyawo, E. S. (2026).
  Distributional Effects with Two-Sided Measurement Error: An Application to
  Intergenerational Income Mobility. arXiv:2107.09235.

- Hausman, J., Liu, H., Luo, Y. and Palmer, C. (2021). Errors in the
  dependent variable of quantile regression models. *Econometrica*, 89(2),
  pp. 849-873.
