---
title: "Two-Sided Measurement Error: Intergenerational Income Mobility"
author: "Brantly Callaway, Tong Li, Irina Murtazashvili, Emmanuel S. Tsyawo"
date: today
format: html
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    fig.width: 7
    fig.height: 4.5
vignette: >
  %\VignetteIndexEntry{Two-Sided Measurement Error: Intergenerational Income Mobility}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

This vignette illustrates `tsme()`, the two-sided measurement error estimator
in the **qrme** package, using the intergenerational income mobility application
from Callaway, Li, Murtazashvili, and Tsyawo (2026).  The key idea is that
when *both* the outcome (child income) and a continuous treatment (father's
income) are measured with error, standard quantile regression estimates of the
conditional distribution of child income given father's income are biased.
`tsme()` corrects this bias by:

1. Running `qrme()` separately for the child income and father income equations
   to recover ME-corrected conditional quantile functions.
2. Linking the two via a copula to recover the joint distribution of latent
   (true) child and father incomes.
3. Deriving mobility statistics from the corrected joint distribution.

## The NLSY97 data

The `nlsy97` dataset contains 1,066 father-son pairs from the National
Longitudinal Survey of Youth 1997.  Key variables:

| Variable | Description |
|---|---|
| `lci` | Log son's annual income |
| `lpi` | Log father's annual income |
| `ageC_97` | Son's age in 1997 |
| `ageF` | Father's age |
| `educF` | Father's years of schooling |

```{r}
#| label: data
library(qrme)
library(ggplot2)

dim(nlsy97)
summary(nlsy97[, c("lci", "lpi", "ageC_97", "ageF", "educF")])
```

Raw (observed) incomes are measured with substantial error — NLSY97 income
data are self-reported and subject to recall error, rounding, and
misclassification.

## Fitting the model with `tsme()`

The call below reproduces the main specification from the paper: a Frank copula,
Laplace measurement error distribution, and one mixture component for each
equation.  Age controls enter both the child income and father income equations.

```{r}
#| label: tsme-call
#| eval: false
tau     <- seq(0.02, 0.98, length.out = 25)
t_values <- quantile(nlsy97$lpi, probs = seq(0.1, 0.9, by = 0.1))

set.seed(51526)
fit <- tsme(
  data            = nlsy97,
  y_formula       = lci ~ ageC_97 + ageF,
  t_formula       = lpi ~ ageC_97 + ageF,
  tau             = tau,
  t_values        = t_values,
  y_n_mix         = 1L,
  t_n_mix         = 1L,
  copula          = "frank",
  me_distribution = "laplace",
  conv_patience   = 1L,
  n_copula_me_draws = 1000L,
  mcmc_draws      = 400L,
  mcmc_burn_in    = 20L,
  conv_criterion  = "params",
  tol             = 1e-2,
  max_em_iters    = 100L,
  se              = TRUE,
  n_boot          = 200L,
  n_cores         = 10L,
  boot_n_copula_me_draws = 500L,
  boot_mcmc_draws = 200L,
  boot_mcmc_burn_in = 10L,
  boot_tol        = 5e-2,
  verbose         = 2
)
```

This call takes roughly 90 minutes with 10 cores.  For the remainder of this
vignette we load the pre-computed result bundled with the package:

```{r}
#| label: load-fit
fit <- nlsy97_tsme_fit
```

## Printing the result

`print()` shows the copula parameter and measurement error distribution for
each equation, comparing the ME-corrected estimator against a no-ME benchmark:

```{r}
#| label: print
print(fit)
```

`summary()` adds the ME-corrected transition matrix and upward mobility
estimates:

```{r}
#| label: summary
summary(fit)
```

## Measurement error parameters

The EM algorithm estimates the ME standard deviation for each equation.  With
a Laplace distribution and one component, the key parameter is $\sigma_V$:

```{r}
#| label: me-params
cat(sprintf(
  "Child income equation  — sigma_V = %.4f  (SE = %.4f)\n",
  fit$me_qyx$sig, fit$me_qyx$sig_se
))
cat(sprintf(
  "Father income equation — sigma_V = %.4f  (SE = %.4f)\n",
  fit$me_qtx$sig, fit$me_qtx$sig_se
))
```

Both equations show substantial measurement error: the ME standard deviation
is roughly 0.50 log-income units (~50% in levels), indicating that
self-reported income data are quite noisy.

## Copula and rank correlation

The Frank copula parameter and the implied Spearman rank correlation summarise
intergenerational dependence:

```{r}
#| label: copula
cat(sprintf(
  "Frank copula parameter: ME = %.3f (SE = %.3f)   No-ME = %.3f\n",
  fit$me_cop_param, fit$me_cop_param_se, fit$nome_cop_param
))
cat(sprintf(
  "Spearman rho:           ME = %.4f (SE = %.4f)   No-ME = %.4f   Observed = %.4f\n",
  fit$me_spearman, fit$me_spearman_se,
  fit$nome_spearman, fit$obs_spearman
))
```

The ME-corrected Spearman rank correlation (0.37) is substantially larger than
the naive observed correlation (0.21), indicating that measurement error causes
a significant downward bias in the apparent degree of intergenerational
persistence.

## Transition matrix

The 4×4 transition matrix gives the probability that a son from father's income
quartile $j$ ends up in son's income quartile $i$:

```{r}
#| label: tmat
quart_labels <- c("Q1 (bottom)", "Q2", "Q3", "Q4 (top)")
tmat_tbl <- as.data.frame(round(fit$me_tmat, 3))
colnames(tmat_tbl) <- quart_labels
rownames(tmat_tbl) <- quart_labels
print(tmat_tbl)
```

The diagonal of the ME-corrected matrix is flatter than the naive observed
matrix, but the off-diagonal persistence is still strong — sons of high-income
fathers are substantially more likely to end up in the top quartile.

## Upward mobility

Upward mobility measures the fraction of sons whose income rank exceeds their
father's, by father's income quartile:

```{r}
#| label: upward-mob
um_tbl <- data.frame(
  `Father's quartile` = quart_labels,
  `ME-corrected`  = round(fit$me_up_mob,    3),
  `No-ME`         = round(fit$nome_up_mob,  3),
  `Observed`      = round(fit$obs_up_mob,   3),
  check.names = FALSE
)
print(um_tbl, row.names = FALSE)
```

Sons of fathers in the bottom income quartile have the highest probability of
upward mobility, while sons of top-quartile fathers have the lowest.

## Conditional quantile curves

`autoplot()` visualises how the conditional distribution of son's income varies
across father's income deciles.  The ME-corrected curves spread further than
the naive QR curves, reflecting the attenuation bias that measurement error
induces in standard QR:

```{r}
#| label: plot-cond-quant
#| fig-cap: "ME-corrected conditional quantile curves"
autoplot(fit, type = "cond_quant", which = "me")
```

```{r}
#| label: plot-all
#| fig-cap: "Conditional quantile curves: ME-corrected vs No-ME vs QR"
# Plot each estimator separately and compare
p_nome <- autoplot(fit, type = "cond_quant", which = "nome", ci = FALSE)
p_qr   <- autoplot(fit, type = "cond_quant", which = "qr",   ci = FALSE)
```

## Poverty rates

The poverty rate is the fraction of sons below a given income threshold,
conditional on father's income decile.  Setting `type = "pov_rate"` and
`which = "all"` overlays all three estimators:

```{r}
#| label: plot-pov
#| fig-cap: "Poverty rates by father's income decile"
autoplot(fit, type = "pov_rate", which = "all", ci = FALSE)
```

## Model selection

`tsme_model_select()` performs a grid search over copula families
(Gaussian, Clayton, Gumbel, Frank) and ME distributions (Gaussian, Laplace)
and returns AIC/BIC for each combination.  This is computationally expensive;
a typical call looks like:

```{r}
#| label: model-select
#| eval: false
sel <- tsme_model_select(
  data            = nlsy97,
  y_formula       = lci ~ ageC_97 + ageF,
  t_formula       = lpi ~ ageC_97 + ageF,
  tau             = tau,
  t_values        = t_values,
  me_distribution = "laplace",  # or c("gaussian", "laplace") for full grid
  mcmc_draws      = 400L,
  mcmc_burn_in    = 20L,
  n_cores         = 10L
)
print(sel)
```

## References

Callaway, B., Li, T., Murtazashvili, I., and Tsyawo, E. S. (2026).
*Distributional Effects with Two-Sided Measurement Error: An Application to
Intergenerational Income Mobility.* arXiv:2107.09235.
[doi:10.48550/arXiv.2107.09235](https://doi.org/10.48550/arXiv.2107.09235)
